---
title: "About"
author: "Brittany Alexandra"
date: "`r Sys.Date()`"

format:
  html:
    theme: cerulean
    toc: true
    toc-depth: 2          
    toc-location: left    
    code-fold: true
    code-tools: true

editor_options: 
  markdown: 
    wrap: 72
---

# **I appreciate you dropping by** 😀

My name's Bri. I enjoy statistical modelling and making predictions.

I graduated with a degree in Statistics from the University of Auckland,
have experienced work as a Data Scientist at New Zealands fastest
growing start-up, as a Research Analyst at one of Vault's top 50
consulting firms, and a Data Analyst at one of top 100 global
universities. All in all, I'm grateful to have such a diverse experience
that shapes my knowledge now. I love learning, and I love sharing, so I
hope you enjoy the content on this website!

This website contains short information bits on statistical modelling
distributions, and examples on how to use them. I hope you find them
useful, and if nothing else- interesting! I sure did.

```{r setup, include = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Install and load required packages

```

```{r}
# Create a data frame for the smiley face
smiley_data <- data.frame(
  x = c(-0.2, 0.2, -0.1, 0.1, -0.1, 0.1),  # X-coordinates
  y = c(0.2, 0.2, 0, 0, -0.2, -0.2)       # Y-coordinates
)

# Create the base plot
p <- ggplot(smiley_data, aes(x, y)) +
  xlim(-1, 1) + ylim(-1, 1) +  # Set the plot limits
  theme_void()                  # Remove background and axis

# Add the face
p <- p +
 
  geom_point(aes(x = -0.3, y = 0.4), size = 10, color = "black") +
  geom_point(aes(x = 0.3, y = 0.4), size = 10, color = "black")

# Add the nose
p <- p +
  geom_point(aes(x = 0, y = 0.), size = 10, color = "red")

# Add the smile
p <- p +
  geom_curve(
    aes(x = -0.4, y = -0.6, xend = 0.4, yend = -0.6),
    color = "black", size = 2, curvature = 0.4
  )

 

# Display the plot
print(p)
```
