---
title: "Introduction to Poisson Modelling"
author: "Brittany Alexandra"
date: "`r Sys.Date()`"

format:
  html:
    theme: cerulean
    toc: true
    toc-depth: 2          
    toc-location: left    
    code-fold: true
    code-tools: true

editor_options: 
  markdown: 
    wrap: 72
---

# I. Modelling processes that evolve randomly over time

------------------------------------------------------------------------

```{r, include= FALSE, warning= FALSE, echo= FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Welcome to the soft introduction into Poisson modelling in R!

Many things in life can be modeled with the poisson process as it quite
literally models the long-term average rate of which random events occur
independently. Its subjectively the most widely used model of a point
process in time, and is suitable for modelling *frequencies* and
probabilities*.*

<br>

[Examples of events that can be modeled by the poisson
distribution]{.underline}

-   The frequency of goals scored by a team in a soccer match

-   The probability two teams will score no goals in a soccer match

-   The frequency of inbound calls to a call center

-   The probability you receive 10 calls in one hour to a call center

-   The frequency of unique visitors to a website

-   The probability you receive 5 unique visitors to a website in 30
    minutes

-   The frequency walk-in arrivals to a physical location (customers or
    patients)

-   The probability you receive 30 walk-in arrivals to a physical
    location in 2 hours.

-   The frequency of volcanic eruptions

-   The probability there will be 2 volcanic eruptions in the next 10
    years.

<br>

[All of these scenarios have the following in common which allow us to
describe the response:]{.underline}

-   Response is discrete

-   Response is right skewed (but approaches a normal distribution as mu
    increases

-   Bounded between 1 and infinity.

-   Variance increases with the mean.

<br>

# II. Synthesis of the poisson model

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

1.  **Poisson Log Link**: Ensures predicted values are +ve.

2.  **Logistic Logit Link**: Logit is between -inf to inf. exp() ensures
    predicted values are bound by 0 and 1.
:::

Now that we've introduced what kinds of real-world phenomena follow a
Poisson distribution, from soccer goals to volcanic eruptions, you might
be wondering, "How do we actually use this knowledge to build a
statistical model?"

In classical linear regression, we assume the response has an
**unbounded continuous range**. That works fine when you're predicting,
say, someone's income or weight. But in **Poisson regression**, we model
**counts**... things like number of visitors, arrivals, or events,
situations where negative values make no sense.

So, instead of the standard form: $\mu_i = \beta_0 + \beta_1 x_i$, we
apply a **log** **link function** to ensure our expected value, $\mu_i$,
is always **positive**. After all, a negative number of customer
walk-ins? That might get you fired 😅

<br>

[**🔗 Enter the Log Link**]{.underline}

To fix this, the Poisson GLM uses a \*\*log link function\*\*, which
transforms the expected value onto the real number line:

$\log(\mu_i) = \beta_0 + \beta_1 x_i \quad \Rightarrow \quad \mu_i = \exp(\beta_0 + \beta_1 x_i)$

[This transformation has two key benefits:]{.underline}

-   \- $\mu_i$ can never be negative

-   \- The model remains linear \*\*on the log scale\*\*

And because of the log transformation, we interpret coefficients
**multiplicatively**...

<br>

The two plots below help visualize how the Poisson model transforms
values using a **log link**. The log ensures the model stays linear on
the log scale, while the exponential ensures predictions for the mean
response, μ, are always positive.

The **log link function** transforms expected counts $\mu$ (which must
be ≥ 0) into linear predictors η on the real number line.

```{r}
library(tidyverse)
 
 # Plot 1: Log Link Function η = log(μ)
mu <- seq(0.01, 10, length.out = 500)
eta_from_mu <- log(mu)
df1 <- data.frame(mu = mu, eta = eta_from_mu)

# Add a point to highlight transformation
highlight1 <- data.frame(mu = 5, eta = log(5))

ggplot(df1, aes(x = mu, y = eta)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(data = highlight1, aes(x = mu, y = eta), color = "red", size = 3) +
  geom_segment(
    data = highlight1,
    aes(x = mu, xend = mu, y = 0, yend = eta),
    linetype = "dashed", color = "gray"
  ) +
  geom_segment(
    data = highlight1,
    aes(x = 0, xend = mu, y = eta, yend = eta),
    linetype = "dashed", color = "gray"
  ) +
  geom_text(
    data = highlight1,
    aes(x = mu + 1.5, y = eta),
    label = expression(paste("η = log(", mu, ") = ", log(5))),
    hjust = 0, size = 5
  ) +
  labs(
    title = "Log Link Function: η = log(μ)",
    x = expression(mu ~ "(mean of response)"),
    y = expression(eta ~ "(linear predictor)")
  ) +
  theme_minimal(base_size = 14)
```

Conversely, The **inverse link function** (exp) transforms any linear
predictor η back into a valid, positive expected count.

```{r}
# Plot 2: Inverse link function (exp)
eta <- seq(-5, 5, length.out = 500)
mu_from_eta <- exp(eta)
df2 <- data.frame(eta = eta, mu = mu_from_eta)

# Highlight a point
highlight2 <- data.frame(eta = 1.6, mu = exp(1.6))

ggplot(df2, aes(x = eta, y = mu)) +
  geom_line(color = "forestgreen", linewidth = 1) +
  geom_point(data = highlight2, aes(x = eta, y = mu), color = "red", size = 3) +
  geom_segment(
    data = highlight2,
    aes(x = eta, xend = eta, y = 0, yend = mu),
    linetype = "dashed", color = "gray"
  ) +
  geom_segment(
    data = highlight2,
    aes(x = 0, xend = eta, y = mu, yend = mu),
    linetype = "dashed", color = "gray"
  ) +
  geom_text(
    data = highlight2,
    aes(x = eta + 0.5, y = mu),
    label = expression(paste("μ = exp(", eta, ") = ", round(exp(1.6), 1))),
    hjust = 0, size = 5
  ) +
  labs(
    title = "Inverse Link Function: μ = exp(η)",
    x = expression(eta ~ "(linear predictor)"),
    y = expression(mu ~ "(mean of response)")
  ) +
  theme_minimal(base_size = 14)
```

# III. Explanatory Terms, and Interpretation

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

Holding all over variables constant, for every n-unit increase in X, the
expected value of the response in multiplied by exp(nB1)
:::

$\log(\mu) = \beta_0 + \beta_1 x$

$\mu = e^{\beta_0 + \beta_1x}$

$=e^{\beta_0} (e^{\beta_1})^x$

\- When $x = 0$, the expected value equals $\exp(\beta_0)$.

\- For every **one-unit** increase in $x$, the expected value of the
response is multiplied by $\exp(\beta_1)$.

\- For every **ten-unit** increase in $x$, the expected value of the
response is multiplied by $\exp(10 \beta_1)$.

\- For every $n$**-unit** increase in $x$, the expected value of the
response is multiplied by $\exp(n \beta_1)$.

# IV. Parameter Estimation for Generalised Linear Models

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

GLMs: Maximise the log-likelihood
:::

**Maximum Log-Likelihood**

As the LSE does not accommodate for increases in variance (and therefore
needs to adjust weighting for the residuals), we use the maximum
likelihood function. We aim to choose coefficients that provide the
highest possible log-likelihood.

[To calculate the likelihood function...]{.underline}

1.  Calculate the probability of observing the observations response

2.  Take the product of these probabilities

    $\ell = \log(L) = \sum_{i=1}^{n} \log \left[ f(y_i; \boldsymbol{\beta}) \right], \ where \ f(y_i; \boldsymbol{\beta}) \ is \ the \ pmf \ of \ the \ assumed \ response \ distribution$

# V. Deviance as a Goodness of Fit

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

1.  Parameter estimation is achieved by minimising the deviance

2.  The more variance, the larger the deviance.

3.  Increasing model complexity always decreases RSS (LR) / increases
    MLE (GLM)

4.  Ds ≤ D ≤ Dn

5.  1 - pchisq(residuals, df) should be large if our model is correct.
:::
