---
title: "Introduction to Poisson Modelling"
author: "Brittany Alexandra"
date: "`r Sys.Date()`"

format:
  html:
    theme: cerulean
    toc: true
    toc-depth: 2          
    toc-location: left    
    code-fold: true
    code-tools: true

editor_options: 
  markdown: 
    wrap: 72
---

# I. Modelling processes that evolve randomly over time

------------------------------------------------------------------------

```{r, include= FALSE, warning= FALSE, echo= FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Welcome to the soft introduction into Poisson modelling in R!

Many things in life can be modeled with the poisson process as it quite
literally models the long-term average rate of which random events occur
independently. Its subjectively the most widely used model of a point
process in time, and is suitable for modelling *frequencies* and
*probabilities.*

<br>

[Examples of events that can be modeled by the poisson
distribution]{.underline}

-   The frequency of goals scored by a team in a soccer match

-   The probability two teams will score no goals in a soccer match

-   The frequency of inbound calls to a call center

-   The probability you receive 10 calls in one hour to a call center

-   The frequency of unique visitors to a website

-   The probability you receive 5 unique visitors to a website in 30
    minutes

-   The frequency walk-in arrivals to a physical location (customers or
    patients)

-   The probability you receive 30 walk-in arrivals to a physical
    location in 2 hours.

-   The frequency of volcanic eruptions

-   The probability there will be 2 volcanic eruptions in the next 10
    years.

<br>

[All of these scenarios have the following in common which allow us to
describe the response:]{.underline}

-   Response is discrete

-   Response is right skewed (but approaches a normal distribution as mu
    increases

-   Bounded between 1 and infinity.

-   Variance increases with the mean.

<br>

[When modelling with the any distribution in R, we will follow the
general format of]{.underline}

1.  Choose a distribution for the response: Poisson

2.  Choose a parameter to relate to explanatory terms: mu_i

3.  Choose a link function: log

4.   Choose explanatory terms (more on this later):

5.  Additional parameters may be estimated/observed:

<br>

# II. Synthesis of the poisson model

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

1.  **Poisson Log Link**: Ensures predicted values are +ve.

2.  **Logistic Logit Link**: Logit is between -inf to inf. exp() ensures
    predicted values are bound by 0 and 1.
:::

Now that we've introduced what kinds of real-world phenomena follow a
Poisson distribution, from soccer goals to volcanic eruptions, you might
be wondering, "How do we actually use this knowledge to build a
statistical model?"

In classical linear regression, we assume the response has an
**unbounded continuous range**. That works fine when you're predicting,
say, someone's income or weight. But in **Poisson regression**, we model
**counts**... things like number of visitors, arrivals, or events,
situations where negative values make no sense.

So, instead of the standard form: $\mu_i = \beta_0 + \beta_1 x_i$, we
apply a **log** **link function** to ensure our expected value, $\mu_i$,
is always **positive**. After all, a negative number of customer
walk-ins? That might get you fired 😅

<br>

[**🔗 Enter the Log Link**]{.underline}

To fix this, the Poisson GLM uses a \*\*log link function\*\*, which
transforms the expected value onto the real number line:

$\log(\mu_i) = \beta_0 + \beta_1 x_i \quad \Rightarrow \quad \mu_i = \exp(\beta_0 + \beta_1 x_i)$

[This transformation has two key benefits:]{.underline}

-   \- $\mu_i$ can never be negative

-   \- The model remains linear \*\*on the log scale\*\*

And because of the log transformation, we interpret coefficients
**multiplicatively**...

<br>

The two plots below help visualize how the Poisson model transforms
values using a **log link**. The log ensures the model stays linear on
the log scale, while the exponential ensures predictions for the mean
response, μ, are always positive.

The **log link function** transforms expected counts $\mu$ (which must
be ≥ 0) into linear predictors η on the real number line.

```{r}
library(tidyverse)
 
 # Plot 1: Log Link Function η = log(μ)
mu <- seq(0.01, 10, length.out = 500)
eta_from_mu <- log(mu)
df1 <- data.frame(mu = mu, eta = eta_from_mu)

# Add a point to highlight transformation
highlight1 <- data.frame(mu = 5, eta = log(5))

ggplot(df1, aes(x = mu, y = eta)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(data = highlight1, aes(x = mu, y = eta), color = "red", size = 3) +
  geom_segment(
    data = highlight1,
    aes(x = mu, xend = mu, y = 0, yend = eta),
    linetype = "dashed", color = "gray"
  ) +
  geom_segment(
    data = highlight1,
    aes(x = 0, xend = mu, y = eta, yend = eta),
    linetype = "dashed", color = "gray"
  ) +
  geom_text(
    data = highlight1,
    aes(x = mu + 1.5, y = eta),
    label = expression(paste("η = log(", mu, ") = ", log(5))),
    hjust = 0, size = 5
  ) +
  labs(
    title = "Log Link Function: η = log(μ)",
    x = expression(mu ~ "(mean of response)"),
    y = expression(eta ~ "(linear predictor)")
  ) +
  theme_minimal(base_size = 14)
```

Conversely, The **inverse link function** (exp) transforms any linear
predictor η back into a valid, positive expected count.

```{r}
# Plot 2: Inverse link function (exp)
eta <- seq(-5, 5, length.out = 500)
mu_from_eta <- exp(eta)
df2 <- data.frame(eta = eta, mu = mu_from_eta)

# Highlight a point
highlight2 <- data.frame(eta = 1.6, mu = exp(1.6))

ggplot(df2, aes(x = eta, y = mu)) +
  geom_line(color = "forestgreen", linewidth = 1) +
  geom_point(data = highlight2, aes(x = eta, y = mu), color = "red", size = 3) +
  geom_segment(
    data = highlight2,
    aes(x = eta, xend = eta, y = 0, yend = mu),
    linetype = "dashed", color = "gray"
  ) +
  geom_segment(
    data = highlight2,
    aes(x = 0, xend = eta, y = mu, yend = mu),
    linetype = "dashed", color = "gray"
  ) +
  geom_text(
    data = highlight2,
    aes(x = eta + 0.5, y = mu),
    label = expression(paste("μ = exp(", eta, ") = ", round(exp(1.6), 1))),
    hjust = 0, size = 5
  ) +
  labs(
    title = "Inverse Link Function: μ = exp(η)",
    x = expression(eta ~ "(linear predictor)"),
    y = expression(mu ~ "(mean of response)")
  ) +
  theme_minimal(base_size = 14)
```

# III. Explanatory Terms, and Interpretation

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

Holding all over variables constant, for every n-unit increase in X, the
expected value of the response in multiplied by exp(B1)
:::

$\log(\mu) = \beta_0 + \beta_1 x$

$\mu = e^{\beta_0 + \beta_1x}$

$=e^{\beta_0} (e^{\beta_1})^x$

\- When $x = 0$, the expected value equals $\exp(\beta_0)$.

\- For every **one-unit** increase in $x$, the expected value of the
response is multiplied by $\exp(\beta_1)$.

\- For every **ten-unit** increase in $x$, the expected value of the
response is multiplied by $\exp(10 \beta_1)$.

\- For every $n$**-unit** increase in $x$, the expected value of the
response is multiplied by $\exp(n \beta_1)$.

# IV. Parameter Estimation for Generalised Linear Models

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

GLMs: Maximise the log-likelihood
:::

**Maximum Log-Likelihood**

In Generalised Linear Models (GLMs), we estimate the model\'s
coefficients using a method called maximum likelihood estimation (MLE).
The core idea is this: for any given set of candidate parameters, we can
compute how likely it is that we would have observed our data, assuming
those parameters are true. This is the **likelihood**.

-   If a data point lies close to the predicted curve (or line), it
    contributes a high likelihood---it\'s consistent with the model.

-   If it\'s far from the predicted values, it contributes a low
    likelihood---it's inconsistent with the model.

<br>
MLE works by searching for the set of parameters (usually denoted as
$\beta$) that maximise the likelihood function. In other words, we
choose the parameters that make our observed data as likely as possible
under the model.
So, instead of minimising the residual sum of squares like in linear
regression, we are maximising the log-likelihood in GLMs. This is
particularly useful when dealing with data where assumptions of constant
variance or normality don\'t hold (e.g. count or binary data).
A better-fitting model will have a higher log-likelihood, meaning it's
more consistent with the observed data.

::: callout-note
Even though two points might look equally close to the predicted line,
their likelihood contributions can still differ due to how spread out
the underlying distribution is. In Poisson regression, this spread grows
with the mean, so deviations are penalized more harshly when the
expected value is small.
:::

```{r}

set.seed(1)
x <- 1:20
y <- rpois(20, lambda = exp(0.2 * x))
model <- glm(y ~ x, family = poisson())
pred <- predict(model, type = "response")
log_lik <- dpois(y, lambda = pred, log = TRUE)

df <- data.frame(x, y, pred, log_lik)

ggplot(df, aes(x, y)) +
  geom_point(aes(color = log_lik), size = 5) +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed") +
  scale_color_viridis_c(option = "virdis") +
  labs(title = "Log-Likelihood Contributions of Points",
       subtitle = "Darker points contribute less to the log-likelihood",
       color = "Log-Likelihood") +
  theme_minimal()
 
```

[To calculate the likelihood function...]{.underline}

1.  Calculate the probability of observing the observations response

2.  Take the product of these probabilities

    $\ell = \log(L) = \sum_{i=1}^{n} \log \left[ f(y_i; \boldsymbol{\beta}) \right], \ where \ f(y_i; \boldsymbol{\beta}) \ is \ the \ pmf \ of \ the \ assumed \ response \ distribution$

# V. Deviance as a Goodness of Fit

------------------------------------------------------------------------

::: callout-note
## 💡 Key Takeaways:

1.  Parameter estimation is achieved by minimising the deviance

2.  The more variance, the larger the deviance.

3.  Increasing model complexity always decreases RSS (LR) / increases
    MLE (GLM)

4.  Ds ≤ D ≤ Dn

5.  1 - pchisq(residuals, df) should be large if our model is correct.
:::

Statistical analysis does not simply end once we have fit a model. We
must determine whether or not our model seems appropriate.
"Goodness-of-fit" is a property that describes how well the data appears
to fit the model's assumptions. For a GLM, these assumptions are

1.  The observations are independent.

2.  $g (\theta) = X \beta$ after applying the link function, the
    parameter of interest is a linear combination of the explanatory
    terms.

3.  Each response comes from the assumed distribution.

<br>

When considering a simple GLM, we can add all sorts of explanatory
terms;

-   Additional numeric explanatory variables

-   Factors via dummy variables

-   Non linear relationships

-   All of the above

<br>

But, adding explanatory terms is sometimes said to make it more
"complex", or increase its "complexity." In consequence, increasing a
models complexity will almost always

-   Decrease the residual sum of squares (linear regression), OR

-   Increase the maximised log-likelihood (GLMS).

**Model Saturation**

```{r}
library(ggplot2)
library(gridExtra)
library(splines)

# Simulate more nonlinear data
set.seed(42)
x <- seq(0, 10, length.out = 10)
y <- 0.5 * x^2 - 3 * x + 5 + rnorm(10, mean = 0, sd = 2)
df <- data.frame(x = x, y = y)

# 1. Underfit model: simple linear regression
model_lm <- lm(y ~ x, data = df)
df$y_lm <- predict(model_lm)

# 2. Moderately saturated: smooth spline with lower spar (more flexible)
spline_fit <- smooth.spline(df$x, df$y, spar = 0.4)  # lower spar for more "wiggle"
x_dense <- seq(min(x), max(x), length.out = 200)
spline_pred <- predict(spline_fit, x_dense)
df_spline <- data.frame(x = spline_pred$x, y = spline_pred$y)

# 3. Fully saturated: perfect interpolation
df$y_sat <- df$y  # fitted values equal observed values

# Plot 1: Underfit (Linear Regression)
p1 <- ggplot(df, aes(x, y)) +
  geom_point(color = "blue") +
  geom_line(aes(y = y_lm), color = "black") +
  geom_segment(aes(xend = x, yend = y_lm), color = "red", linetype = "dashed") +
  ggtitle("Underfit Model") +
  theme_minimal()

# Plot 2: Moderately Saturated (Spline Fit)
p2 <- ggplot(df, aes(x, y)) +
  geom_point(color = "blue") +
  geom_line(data = df_spline, aes(x = x, y = y), color = "orange") +
  geom_segment(aes(xend = x, yend = predict(spline_fit, x)$y), color = "red", linetype = "dashed") +
  ggtitle("Moderately Saturated Model") +
  theme_minimal()

# Plot 3: Fully Saturated (Perfect Fit)
p3 <- ggplot(df, aes(x, y)) +
  geom_point(color = "blue") +
  geom_line(aes(y = y_sat), color = "green") +
  ggtitle("Fully Saturated Model") +
  theme_minimal()

# Combine the three plots
grid.arrange(p1, p2, p3, nrow = 1)


```

In almost all cases, the saturated model is not a sensible model as

1.  If we were to take another sample from the population, their
    observations would not exhibit the exact same trend.

2.  The exact pattern we see therefore reflects the random white noise
    inherent in our particular sample, and not the underlying
    relationship that exists in the population.

3.  We should try avoid over fitting models, such as the saturated
    models, so that they only describe the underlying trends in the
    population, and not random noise from our sample
